---
title: 'Foundations for statistical inference - Confidence intervals'
author: "Mehreen Ali Gillani"
output:
  pdf_document: default
  html_document:
    includes:
      in_header: header.html
    css: ./lab.css
    highlight: pygments
    theme: cerulean
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(eval = TRUE, message = FALSE, warning = FALSE)
set.seed(123)
```

If you have access to data on an entire population, say the opinion of every adult in the United States on whether or not they think climate change is affecting their local community, it's straightforward to answer questions like, "What percent of US adults think climate change is affecting their local community?". Similarly, if you had demographic information on the population you could examine how, if at all, this opinion varies among young and old adults and adults with different leanings. If you have access to only a sample of the population, as is often the case, the task becomes more complicated. What is your best guess for this proportion if you only have data from a small sample of adults? This type of situation requires that you use your sample to make inference on what your population looks like.

<div id="boxedtext">
**Setting a seed:** You will take random samples and build sampling distributions in this lab, which means you should set a seed on top of your lab. If this concept is new to you, review the lab on probability.
</div>

## Getting Started

### Load packages

In this lab, we will explore and visualize the data using the **tidyverse** suite of packages, and perform statistical inference using **infer**.

Let's load the packages.

```{r load-packages, message=FALSE}
library(tidyverse)
library(openintro)
library(infer)
```


### The data

A 2019 Pew Research report states the following:

To keep our computation simple, we will assume a total population size of 100,000 (even though that's smaller than the population size of all US adults).

> Roughly six-in-ten U.S. adults (62%) say climate change is currently affecting their local community either a great deal or some, according to a new Pew Research Center survey.
>
>**Source:** [Most Americans say climate change impacts their community, but effects vary by region](https://www.pewresearch.org/fact-tank/2019/12/02/most-americans-say-climate-change-impacts-their-community-but-effects-vary-by-region/)

In this lab, you will assume this 62% is a true population proportion and learn about how sample proportions can vary from sample to sample by taking smaller samples from the population. We will first create our population assuming a population size of 100,000. This means 62,000 (62%) of the adult population think climate change impacts their community, and the remaining 38,000 does not think so.

```{r}
us_adults <- tibble(
  climate_change_affects = c(rep("Yes", 62000), rep("No", 38000))
)
```

The name of the data frame is `us_adults` and the name of the variable that contains responses to the question *"Do you think climate change is affecting your local community?"* is `climate_change_affects`.

We can quickly visualize the distribution of these responses using a bar plot.

```{r bar-plot-pop, fig.height=2.5, fig.width=10}
ggplot(us_adults, aes(x = climate_change_affects)) +
  geom_bar() +
  labs(
    x = "", y = "",
    title = "Do you think climate change is affecting your local community?"
  ) +
  coord_flip() 
```

We can also obtain summary statistics to confirm we constructed the data frame correctly.

```{r summ-stat-pop, results = TRUE}
us_adults %>%
  count(climate_change_affects) %>%
  mutate(p = n /sum(n))
```

In this lab, you'll start with a simple random sample of size 60 from the population.

```{r sample}
n <- 60
samp <- us_adults %>%
  sample_n(size = n)
```

1.  What percent of the adults in your sample think climate change affects their 
    local community? **Hint:** Just like we did with the population, we can calculate 
    the proportion of those **in this sample** who think climate change affects their 
    local community.

**Insert your answer here**
```{r}
samp_p_hat <- samp %>% 
  count(climate_change_affects) %>%
  mutate(p_hat = n /sum(n)) %>%
  filter(climate_change_affects=="Yes") 
samp_percentage <- samp_p_hat$p_hat * 100
samp_percentage
```

65% people think climate change affects their local community.

1.  Would you expect another student's sample proportion to be identical to yours? 
    Would you expect it to be similar? Why or why not?

**Insert your answer here**

I have set a seed so running same code above will give me same answer. But let me take another sample of size 60  and find its percentage.

```{r}
set.seed(136)
n <- 60
sam <- us_adults %>%
  sample_n(size = n)
samp_p_hat1 <- sam %>% 
  count(climate_change_affects) %>%
  mutate(p_hat1 = n /sum(n)) %>%
  filter(climate_change_affects=="Yes") 
samp_percentage1 <- samp_p_hat1$p_hat1 * 100
samp_percentage1
```

This time it is different 61.667% which means my classmates will find a different result. It makes sense because we are randomly selecting 60 people from the poll. SO everyone will have different result in percentage. 

## Confidence intervals

Return for a moment to the question that first motivated this lab: based on this sample, what can you infer about the population? With just one sample, the best estimate of the proportion of US adults who think climate change affects their local community would be the sample proportion, usually denoted as $\hat{p}$ (here we are calling it `p_hat`). That serves as a good **point estimate**, but it would be useful to also communicate how uncertain you are of that estimate. This uncertainty can be quantified using a **confidence interval**.

One way of calculating a confidence interval for a population proportion is based on the Central Limit Theorem, as $\hat{p} \pm z^\star SE_{\hat{p}}$ is, or more precisely, as \[ \hat{p} \pm z^\star \sqrt{ \frac{\hat{p} (1-\hat{p})}{n} } \]

Another way is using simulation, or to be more specific, using **bootstrapping**. The term **bootstrapping** comes from the phrase "pulling oneself up by one's bootstraps", which is a metaphor for accomplishing an impossible task without any outside help. In this case the impossible task is estimating a population parameter (the unknown population proportion), and we'll accomplish it using data from only the given sample. Note that this notion of saying something about a population parameter using only information from an observed sample is the crux of statistical inference, it is not limited to bootstrapping. 

In essence, bootstrapping assumes that there are more of observations in the populations like the ones in the observed sample. So we "reconstruct" the population by resampling from our sample, with replacement. The bootstrapping scheme is as follows:

- **Step 1.** Take a bootstrap sample - a random sample taken **with replacement** from the original sample, of the same size as the original sample.
- **Step 2.** Calculate the bootstrap statistic - a statistic such as mean, median, proportion, slope, etc. computed on the bootstrap samples.
- **Step 3.** Repeat steps (1) and (2) many times to create a bootstrap distribution - a distribution of bootstrap statistics.
- **Step 4.** Calculate the bounds of the XX% confidence interval as the middle XX% j knof the bootstrap distribution.

Instead of coding up each of these steps, we will construct confidence intervals using the **infer** package.

Below is an overview of the functions we will use to construct this confidence interval:

Function    | Purpose
----------- | -------
`specify`   | Identify your variable of interest
`generate`  | The number of samples you want to generate
`calculate` | The sample statistic you want to do inference with, or you can also think of this as the population parameter you want to do inference for
`get_ci`    | Find the confidence interval

This code will find the 95 percent confidence interval for proportion of US adults who think climate change affects their local community.

```{r confidence interval infer}
samp %>%
  specify(response = climate_change_affects, success = "Yes") %>%
  generate(reps = 1000, type = "bootstrap") %>%
  calculate(stat = "prop") %>%
  get_ci(level = 0.95)
```

- In `specify` we specify the `response` variable and the level of that variable we are calling a `success`.
- In `generate` we provide the number of resamples we want from the population in the `reps` argument (this should be a reasonably large number) as well as the type of resampling we want to do, which is `"bootstrap"` in the case of constructing a confidence interval.
- Then, we `calculate` the sample statistic of interest for each of these resamples, which is `prop`ortion.

Feel free to test out the rest of the arguments for these functions, since these commands will be used together to calculate confidence intervals and solve inference problems for the rest of the semester. But we will also walk you through more examples in future chapters.

To recap: even though we don't know what the full population looks like, we're 95% confident that the true proportion of US adults who think climate change affects their local community is between the two bounds reported as result of this pipeline.

## Confidence levels

1.  In the interpretation above, we used the phrase "95% confident". What does "95% confidence" mean?

**Insert your answer here**


If we were to take many repeated random samples from the same population and construct a 95% confidence interval from each sample, then about 95% of those intervals would contain the true population parameter.

Different samples will produce different intervals, but 95% of them will be "correct"

The process of building intervals works 95% of the time, it is called method reliability. 


In this case, you have the rare luxury of knowing the true population proportion (62%) since you have data on the entire population. 

1.  Does your confidence interval capture the true population proportion of US adults 
    who think climate change affects their local community? If you are working on this 
    lab in a classroom, does your neighbor's interval capture this value? 

**Insert your answer here**
Yes, my confidence interval successfully captures the true population proportion. With a confidence interval ranging from 0.517 to 0.767, the true population proportion of 0.62 falls well within this range.

W each student will obtain slightly different confidence interval bounds due to random sampling variation, the vast majority of their intervals should also contain the true population proportion of 0.62. This is exactly what the 95% confidence level means - that approximately 95% of all possible samples will produce confidence intervals that capture the true population parameter.


1.  Each student should have gotten a slightly different confidence interval. What 
    proportion of those intervals would you expect to capture the true population 
    mean? Why?

**Insert your answer here**

This is the fundamental definition of a 95% confidence level. If every student in the class:

Takes a random sample from the same population
Constructs a 95% confidence interval using the same method
Repeats this process many times

Then about 95% of all the intervals generated would contain the true population parameter. The confidence level describes the long-run performance of the interval construction method, not the probability that any one specific interval contains the parameter.

In the next part of the lab, you will collect many samples to learn more about how sample proportions and confidence intervals constructed based on those samples vary from one sample to another.

-   Obtain a random sample.
-   Calculate the sample proportion, and use these to calculate and store the lower and upper bounds of the confidence intervals.
-   Repeat these steps 50 times.

Doing this would require learning programming concepts like iteration so that you can automate repeating running the code you've developed so far many times to obtain many (50) confidence intervals. In order to keep the programming simpler, we are providing the interactive app below that basically does this for you and created a plot similar to Figure 5.6 on [OpenIntro Statistics, 4th Edition (page 182)](https://www.openintro.org/os).
```{r define-store-ci}
# Define the store_ci function
store_ci <- function(i, n, reps, conf_level, success) {
  us_adults %>%
    sample_n(size = n) %>%
    specify(response = climate_change_affects, success = success) %>%
    generate(reps, type = "bootstrap") %>%
    calculate(stat = "prop") %>%
    get_ci(level = conf_level) %>%
    rename(
      x_lower = names(.)[1],
      x_upper = names(.)[2]
    )
}
```
```{r shiny, echo=FALSE, eval=FALSE, results = TRUE}
# This R chunk will only run in interactive mode
store_ci <- function(i, n, reps, conf_level, success) {
  us_adults %>%
    sample_n(size = n) %>%
    specify(response = climate_change_affects, success = success) %>%
    generate(reps, type = "bootstrap") %>%
    calculate(stat = "prop") %>%
    get_ci(level = conf_level) %>%
    rename(
      x_lower = names(.)[1],
      x_upper = names(.)[2]
    )
}
library(shiny)
shinyApp(
  ui <- fluidPage(
    h4("Confidence intervals for the proportion of US adults who think 
     climate change"),

    h4(selectInput("success", "",
      choices = c(
        "is affecting their local community" = "Yes",
        "is not affecting their local community" = "No"
      ),
      selected = "Yes", width = "50%"
    )),

    # Sidebar with a slider input for number of bins
    sidebarLayout(
      sidebarPanel(
        numericInput("n_samp",
          "Sample size for a single sample from the population:",
          min = 1,
          max = 1000,
          value = 60
        ),

        hr(),

        numericInput("n_rep",
          "Number of resamples for each bootstrap confidence interval:",
          min = 1,
          max = 15000,
          value = 1000
        ),

        numericInput("conf_level",
          "Confidence level",
          min = 0.01,
          max = 0.99,
          value = 0.95,
          step = 0.05
        ),

        hr(),

        radioButtons("n_ci",
          "Number of confidence intervals:",
          choices = c(10, 25, 50, 100),
          selected = 50, inline = TRUE
        ),

        actionButton("go", "Go")
      ),

      # Show a plot of the generated distribution
      mainPanel(
        plotOutput("ci_plot")
      )
    )
  ),

  server <- function(input, output) {

    # set true p
    p <- reactive(ifelse(input$success == "Yes", 0.62, 0.38))

    # create df_ci when go button is pushed
    df_ci <- eventReactive(input$go, {
      map_dfr(1:input$n_ci, store_ci,
        n = input$n_samp,
        reps = input$n_rep, conf_level = input$conf_level,
        success = input$success
      ) %>%
        mutate(
          y_lower = 1:input$n_ci,
          y_upper = 1:input$n_ci,
          capture_p = ifelse(x_lower < p() & x_upper > p(), "Yes", "No")
        )
    })

    # plot df_ci
    output$ci_plot <- renderPlot({
      ggplot(df_ci()) +
        geom_segment(aes(x = x_lower, y = y_lower, xend = x_upper, yend = y_upper, color = capture_p)) +
        geom_point(aes(x = x_lower, y = y_lower, color = capture_p)) +
        geom_point(aes(x = x_upper, y = y_upper, color = capture_p)) +
        geom_vline(xintercept = p(), color = "darkgray") +
        labs(
          y = "", x = "Bounds of the confidence interval",
          color = "Does the interval capture the true population proportion?"
        ) +
        theme(legend.position = "bottom")
    })
  },
  options = list(height = 700)
)
```


1.  Given a sample size of 60, 1000 bootstrap samples for each interval, and 50 
    confidence intervals constructed (the default values for the above app), what 
    proportion of your confidence intervals include the true population proportion? 
    Is this proportion exactly equal to the confidence level? If not, explain why. 
    Make sure to include your plot in your answer.

**Insert your answer here**


```{r bootstrap coverage}
set.seed(123)
library(infer)
library(dplyr)
library(ggplot2)

# Initialize storage
true_prop = 0.62
capture_results <- logical(50)
lower_bounds <- numeric(50)
upper_bounds <- numeric(50)
widths95 <- numeric(50)
for(i in 1:50) {
  ci <- store_ci(i, n = 60, reps = 1000, conf_level = 0.95, success = "Yes")
  
  # Store results - access columns directly
  lower_bounds[i] <- ci$x_lower
  upper_bounds[i] <- ci$x_upper
  widths95 [i] <- ci$x_upper - ci$x_lower
  capture_results[i] <- (true_prop >= ci$x_lower & ci$x_upper >= true_prop)
}

# Calculate results
prop_captured <- mean(capture_results)
num_captured <- sum(capture_results)

cat("Proportion capturing true proportion:", prop_captured, "\n")
cat("Number captured:", num_captured, "out of 50\n")
cat("Expected (95% confidence):", 0.95, "\n")
cat("Width of confidence Interval 95 is",mean(widths95),"\n")
# Create plot data
plot_data <- data.frame(
  interval = 1:50,
  lower = lower_bounds,
  upper = upper_bounds,
  captures_true = factor(capture_results, levels = c(TRUE, FALSE))
)

# Create the plot
ggplot(plot_data, aes(x = interval)) +
  geom_errorbar(aes(ymin = lower, ymax = upper, color = captures_true), 
                width = 0.5, alpha = 0.7) +
  geom_hline(yintercept = true_prop, linetype = "dashed", 
             color = "red", linewidth = 1) +
  labs(title = "50 Bootstrap Confidence Intervals",
       subtitle = paste("Mean of capturing true proportion =", round(prop_captured, 3)),
       x = "Interval Number",
       y = "Proportion") +
  scale_color_manual(values = c("TRUE" = "darkgreen", "FALSE" = "darkred"),
                     name = "Captures True Proportion") +
  theme_minimal()
```

with (95% confidence)

Proportion capturing true proportion: 0.92 

Number captured: 46 out of 50

width is 0.241

* * *

## More Practice

1.  Choose a different confidence level than 95%. Would you expect a confidence 
    interval at this level to be wider or narrower than the confidence interval 
    you calculated at the 95% confidence level? Explain your reasoning.

**Insert your answer here**

I will try different confidence interval. first I will try 90%

```{r bootstrap-coverage-analysis CI 90%}
set.seed(123)


# Initialize storage
capture_results <- logical(50)
lower_bounds <- numeric(50)
upper_bounds <- numeric(50)
widths_90 <- numeric(50)
for(i in 1:50) {
  ci <- store_ci(i, n = 60, reps = 1000, conf_level = 0.90, success = "Yes")
  
  # Store results - access columns directly
  lower_bounds[i] <- ci$x_lower
  upper_bounds[i] <- ci$x_upper
  widths_90[i] <- ci$x_upper - ci$x_lower  # Calculate width
  capture_results[i] <- (true_prop >= ci$x_lower & ci$x_upper >= true_prop)
}

# Calculate results
prop_captured <- mean(capture_results)
num_captured <- sum(capture_results)

cat("Proportion capturing true proportion:", prop_captured, "\n")
cat("Number captured:", num_captured, "out of 50\n")
cat("Expected (90% confidence):", 0.90, "\n")
mean(widths_90)
```

Proportion capturing true proportion: 0.86 

Number captured: 43 out of 50

width of 90 Confidence interval is 0.202

width of CI=90% (0.202) is smaller than width of CI=95% (0.241)


Lets try 99% CI
```{r}
widths_99 <- numeric(50)
for(i in 1:50) {
  ci <- store_ci(i, n = 60, reps = 1000, conf_level = 0.99, success = "Yes")
  
  # Store results - access columns directly
  lower_bounds[i] <- ci$x_lower
  upper_bounds[i] <- ci$x_upper
  widths_99[i] <- ci$x_upper - ci$x_lower  # Calculate width
  capture_results[i] <- (true_prop >= ci$x_lower & ci$x_upper >= true_prop)
}

# Calculate results
prop_captured <- mean(capture_results)
num_captured <- sum(capture_results)

cat("Proportion capturing true proportion:", prop_captured, "\n")
cat("Number captured:", num_captured, "out of 50\n")
cat("Expected (99% confidence):", 0.99, "\n")
mean(widths_99)

```
Proportion capturing true proportion: 1 

Number captured: 50 out of 50

width of CI=95% (0.241) is smaller than width of CI=99% (0.309) 


99% CI: Width = 0.309
95% CI: Width = 0.241
90% CI: Width = 0.202 

99% CI is wider than 95% and 95% is wider than 90% CI


To be more confident that we've captured the true population parameter, we need to cast a wider net. I will choose 99% CI,
if I have to choose other than 95% CI.

1.  Using code from the **infer** package and data fromt the one sample you have 
    (`samp`), find a confidence interval for the proportion 
    of US Adults who think climate change is affecting their local community with a 
    confidence level of your choosing (other than 95%) and interpret it.
    
**Insert your answer here**    
```{r bootstrap-coverage-analysis 99}
capture_results <- logical(50)
lower_bounds <- numeric(50)
upper_bounds <- numeric(50)
widths_99 <- numeric(50)
for(i in 1:50) {
  ci <- store_ci(i, n = 60, reps = 1000, conf_level = 0.99, success = "Yes")
  
  # Store results - access columns directly
  lower_bounds[i] <- ci$x_lower
  upper_bounds[i] <- ci$x_upper
  widths_99 [i] <- ci$x_upper - ci$x_lower
  capture_results[i] <- (true_prop >= ci$x_lower & ci$x_upper >= true_prop)
}
# Calculate results
prop_captured <- mean(capture_results)
num_captured <- sum(capture_results)

# Create plot data
plot_data <- data.frame(
  interval = 1:50,
  lower = lower_bounds,
  upper = upper_bounds,
  captures_true = factor(capture_results, levels = c(TRUE, FALSE))
)
mean(widths_99)
# Create the plot
ggplot(plot_data, aes(x = interval)) +
  geom_errorbar(aes(ymin = lower, ymax = upper, color = captures_true), 
                width = 0.5, alpha = 0.7) +
  geom_hline(yintercept = true_prop, linetype = "dashed", 
             color = "red", linewidth = 1) +
  labs(title = "50 Bootstrap Confidence Intervals 99",
       subtitle = paste("Mean of capturing true proportion =", round(prop_captured, 3)),
       x = "Interval Number",
       y = "Proportion") +
  scale_color_manual(values = c("TRUE" = "darkgreen", "FALSE" = "darkred"),
                     name = "Captures True Proportion") +
  theme_minimal()
```


1.  Using the app, calculate 50 confidence intervals at the confidence level you chose 
    in the previous question, and plot all intervals on one plot, and calculate 
    the proportion of intervals that include the true population proportion. 
    How does this percentage compare to the confidence level selected for the 
    intervals?
    
**Insert your answer here**


For the most part the result of the app for 50 Intervals for 99% confidence interval, 1000 rep is same as shown in graph above 

1.  Lastly, try one more (different) confidence level. First, state how you expect the
    width of this interval to compare to previous ones you calculated. Then, 
    calculate the bounds of the interval using the **infer** package and data 
    from `samp` and interpret it. Finally, use the app to generate many intervals 
    and calculate the proportion of intervals that are capture the true population 
    proportion.

**Insert your answer here**
Lets try confidence interval 80%

```{r bootstrap-coverage-analysis 80}
capture_results <- logical(50)
lower_bounds <- numeric(50)
upper_bounds <- numeric(50)
widths_80 <- numeric(50)
for(i in 1:50) {
  ci <- store_ci(i, n = 60, reps = 1000, conf_level = 0.80, success = "Yes")
  
  # Store results - access columns directly
  lower_bounds[i] <- ci$x_lower
  upper_bounds[i] <- ci$x_upper
  widths_80 [i] <- ci$x_upper - ci$x_lower
  capture_results[i] <- (true_prop >= ci$x_lower & ci$x_upper >= true_prop)
}

# Calculate results
prop_captured <- mean(capture_results)
num_captured <- sum(capture_results)
# Create plot data
plot_data <- data.frame(
  interval = 1:50,
  lower = lower_bounds,
  upper = upper_bounds,
  captures_true = factor(capture_results, levels = c(TRUE, FALSE))
)
mean(widths_80)
# Create the plot
ggplot(plot_data, aes(x = interval)) +
  geom_errorbar(aes(ymin = lower, ymax = upper, color = captures_true), 
                width = 0.5, alpha = 0.7) +
  geom_hline(yintercept = true_prop, linetype = "dashed", 
             color = "red", linewidth = 1) +
  labs(title = "50 Bootstrap Confidence Intervals 80",
       subtitle = paste("Mean of capturing true proportion =", round(prop_captured, 3)),
       x = "Interval Number",
       y = "Proportion") +
  scale_color_manual(values = c("TRUE" = "darkgreen", "FALSE" = "darkred"),
                     name = "Captures True Proportion") +
  theme_minimal()
```


1.  Using the app, experiment with different sample sizes and comment on how the 
    widths of intervals change as sample size changes (increases and decreases).
    
**Insert your answer here**

1.  Finally, given a sample size (say, 60), how does the width of the interval change 
    as you increase the number of bootstrap samples. **Hint:** Does changing the 
    number of bootstap samples affect the standard error?
    
**Insert your answer here**

Lets change different reps =5000 for conf_level=0.80

```{r bootstrap-coverage-analysis 80 5000}
set.seed(123)
capture_results <- logical(50)
lower_bounds <- numeric(50)
upper_bounds <- numeric(50)
widths_80 <- numeric(50)
standard_errors <- numeric(50)  # Store standard errors
for(i in 1:50) {
  # Get the bootstrap distribution first
  bootstrap_dist <- us_adults %>%
    sample_n(size = 60) %>%
    specify(response = climate_change_affects, success = "Yes") %>%
    generate(reps = 5000, type = "bootstrap") %>%
    calculate(stat = "prop")
  
  # Calculate standard error from bootstrap distribution
  standard_errors[i] <- sd(bootstrap_dist$stat)
  
  # Then get confidence interval
  ci <- get_ci(bootstrap_dist, level = 0.80)
  
  # Store results
  lower_bounds[i] <- ci$lower_ci
  upper_bounds[i] <- ci$upper_ci
  widths_80[i] <- ci$upper_ci - ci$lower_ci
  capture_results[i] <- (true_prop >= ci$lower_ci & ci$upper_ci >= true_prop)
}

# Calculate results
prop_captured <- mean(capture_results)
num_captured <- sum(capture_results)
avg_width <- mean(widths_80)
avg_se <- mean(standard_errors)

cat("80% CONFIDENCE INTERVAL RESULTS:\n")
cat("Proportion capturing true proportion:", round(prop_captured, 3), "\n")
cat("Number captured:", num_captured, "out of 50\n")
cat("Average width of 80% CI:", round(avg_width, 4), "\n")
cat("Average standard error:", round(avg_se, 4), "\n")
cat("Expected coverage (80% confidence): 0.80\n")
```

Now lets try 1000 reps

```{r bootstrap-coverage-analysis 80 1000}
set.seed(123)
capture_results <- logical(50)
lower_bounds <- numeric(50)
upper_bounds <- numeric(50)
widths_80 <- numeric(50)
standard_errors <- numeric(50)  # Store standard errors
for(i in 1:50) {
  # Get the bootstrap distribution first
  bootstrap_dist <- us_adults %>%
    sample_n(size = 60) %>%
    specify(response = climate_change_affects, success = "Yes") %>%
    generate(reps = 1000, type = "bootstrap") %>%
    calculate(stat = "prop")
  
  # Calculate standard error from bootstrap distribution
  standard_errors[i] <- sd(bootstrap_dist$stat)
  
  # Then get confidence interval
  ci <- get_ci(bootstrap_dist, level = 0.80)
  
  # Store results
  lower_bounds[i] <- ci$lower_ci
  upper_bounds[i] <- ci$upper_ci
  widths_80[i] <- ci$upper_ci - ci$lower_ci
  capture_results[i] <- (true_prop >= ci$lower_ci & ci$upper_ci >= true_prop)
}

# Calculate results
prop_captured <- mean(capture_results)
num_captured <- sum(capture_results)
avg_width <- mean(widths_80)
avg_se <- mean(standard_errors)

cat("80% CONFIDENCE INTERVAL RESULTS:\n")
cat("Proportion capturing true proportion:", round(prop_captured, 3), "\n")
cat("Number captured:", num_captured, "out of 50\n")
cat("Average width of 80% CI:", round(avg_width, 4), "\n")
cat("Average standard error:", round(avg_se, 4), "\n")
cat("Expected coverage (80% confidence): 0.80\n")
```

for 5000 reps 80% CONFIDENCE INTERVAL RESULTS:

Proportion capturing true proportion: 0.82 

Number captured: 41 out of 50

Average width of 80% CI: 0.164 

Average standard error: 0.0625 


for 1000 reps 80% CONFIDENCE INTERVAL RESULTS:

Proportion capturing true proportion: 0.74 

Number captured: 37 out of 50

Average width of 80% CI: 0.1621 

Average standard error: 0.0622 

As Number of bootstrap increased from 1000 to 5000

Proportion capturing true proportion has increased from 0.74 to 0.82

Average standard error has slightly decreased by 0.0019

Average width has also slightly decreased by 0.003

Law of Large Numbers: Both 1000 and 5000 reps are sufficient for stable variance estimates

Bootstrap Consistency: Standard error estimates stabilize quickly with bootstrap

Trade-off: 5x more computation for marginal improvement in coverage accuracy, with virtually no change in precision metrics.


* * *
