---
title: "Multiple linear regression"
author: "Mehreen Ali Gillani"
output:
  pdf_document: default
  html_document:
    includes:
      in_header: header.html
    css: ./lab.css
    highlight: pygments
    theme: cerulean
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(eval = TRUE, message = FALSE, warning = FALSE)
```

## Grading the professor

Many college courses conclude by giving students the opportunity to evaluate the course and the instructor anonymously. However, the use of these student evaluations as an indicator of course quality and teaching effectiveness is often criticized because these measures may reflect the influence of non-teaching related characteristics, such as the physical appearance of the instructor. The article titled, "Beauty in the classroom: instructors' pulchritude and putative pedagogical productivity" by Hamermesh and Parker found that instructors who are viewed to be better looking receive higher instructional ratings. 

Here, you will analyze the data from this study in order to learn what goes into a positive professor evaluation.

## Getting Started

### Load packages

In this lab, you will explore and visualize the data using the **tidyverse** suite of packages. The data can be found in the companion package for OpenIntro resources, **openintro**.

Let's load the packages.

```{r load-packages, message=FALSE}
library(tidyverse)
library(openintro)
library(GGally)
```

This is the first time we're using the `GGally` package. You will be using the `ggpairs` function from this package later in the lab.

### The data

The data were gathered from end of semester student evaluations for a large sample of professors from the University of Texas at Austin. In addition, six students rated the professors' physical appearance. The result is a data frame where each row contains a different course and columns represent variables about the courses and professors. It's called `evals`.

```{r}
glimpse(evals)
```

We have observations on 21 different variables, some categorical and some numerical. The meaning of each variable can be found by bringing up the help file:

```{r help-evals, eval=FALSE}
?evals
```

## Exploring the data

1.  Is this an observational study or an experiment? The original research
    question posed in the paper is whether beauty leads directly to the
    differences in course evaluations. Given the study design, is it possible to
    answer this question as it is phrased? If not, rephrase the question.

**Insert your answer here**

Observational Study: No intervention; only observation, conclude and show correlation/association	
Experiment: Active intervention with a treatment, Can suggest causation

By the definiation of observational study and Experiment. It is observational study. 
The study is observational, so it can only identify association, not prove that beauty "leads directly to" (causes) differences in evaluations. The causal claim is inappropriate due to potential confounding variables (e.g., a professor's communication skills might influence both their perceived beauty and their ratings).

Rephrased Question:
"Is there an association between a professor's physical appearance and their course evaluation scores?"

2.  Describe the distribution of `score`. Is the distribution skewed? What does 
    that tell you about how students rate courses? Is this what you expected to 
    see? Why, or why not?

**Insert your answer here**

```{r}
# Plot distribution using ggplot2
library(ggplot2)
ggplot(evals, aes(x = score)) +
  geom_histogram(binwidth = 0.2, fill = "lightblue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Professor Evaluation Scores",
       x = "Score",
       y = "Frequency") +
  theme_minimal()

```

  * Scores are typically left-skewed, clustering around 4.0-4.5
  * Few professors receive very low scores (<3.0)


3.  Excluding `score`, select two other variables and describe their relationship 
    with each other using an appropriate visualization.

**Insert your answer here**
```{r}
colnames(evals)
```
Relationship between Professor Age and Beauty Rating
```{r}
library(ggplot2)
ggplot(evals, aes(x = age, y = bty_avg)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Relationship between Professor Age and Beauty Rating",
       x = "Age", 
       y = "Average Beauty Rating") +
  theme_minimal()
```

Relationship: There appears to be a negative correlation between professor age and beauty ratings. As age increases, beauty ratings tend to decrease.

## Simple linear regression

The fundamental phenomenon suggested by the study is that better looking teachers are evaluated more favorably. Let's create a scatterplot to see if this appears to be the case:

```{r scatter-score-bty_avg}
ggplot(data = evals, aes(x = bty_avg, y = score)) +
  geom_point()
```

Before you draw conclusions about the trend, compare the number of observations in the data frame with the approximate number of points on the scatterplot. Is anything awry?

4.  Replot the scatterplot, but this time use `geom_jitter` as your layer. What 
    was misleading about the initial scatterplot?

```{r scatter-score-bty_avg-jitter}
ggplot(data = evals, aes(x = bty_avg, y = score)) +
  geom_jitter()
```

**Insert your answer here**
```{r}
lm(score ~ bty_avg, data = evals) |> summary()

ggplot(evals, aes(bty_avg, score)) + 
  geom_jitter() + 
  geom_smooth(method = "lm")
```

Based on these results:

Weak relationship: The R-squared of 0.035 means that only 3.5% of the variation in professor evaluation scores is explained by beauty ratings.
Statistically significant, but practically weak: The very small p-value (5.083e-05) indicates the relationship is not due to random chance, but the low R-squared shows beauty has minimal practical impact on scores.

Positive correlation: Since the coefficient for bty_avg would be positive, higher beauty ratings are associated with slightly higher evaluation scores, but the effect is very small.



5.  Let's see if the apparent trend in the plot is something more than
    natural variation. Fit a linear model called `m_bty` to predict average
    professor score by average beauty rating. Write out the equation for the linear 
    model and interpret the slope. Is average beauty score a statistically significant
    predictor? Does it appear to be a practically significant predictor?
    
**Insert your answer here**
```{r}
# Fit the linear model
m_bty <- lm(score ~ bty_avg, data = evals)

# Get the equation coefficients
coefficients(m_bty)

# View model summary
summary(m_bty)
```

Add the line of the bet fit model to your plot using the following:
    
```{r scatter-score-bty_avg-line-se}
ggplot(data = evals, aes(x = bty_avg, y = score)) +
  geom_jitter() +
  geom_smooth(method = "lm")
```

The blue line is the model. The shaded gray area around the line tells you about the variability you might expect in your predictions. To turn that off, use `se = FALSE`.

```{r scatter-score-bty_avg-line}
ggplot(data = evals, aes(x = bty_avg, y = score)) +
  geom_jitter() +
  geom_smooth(method = "lm", se = FALSE)
```

6.  Use residual plots to evaluate whether the conditions of least squares
    regression are reasonable. Provide plots and comments for each one (see
    the Simple Regression Lab for a reminder of how to make these).

**Insert your answer here**

## Multiple linear regression

The data set contains several variables on the beauty score of the professor: individual ratings from each of the six students who were asked to score the physical appearance of the professors and the average of these six scores. Let's take a look at the relationship between one of these scores and the average beauty score.

```{r bty-rel}
ggplot(data = evals, aes(x = bty_f1lower, y = bty_avg)) +
  geom_point()

evals %>% 
  summarise(cor(bty_avg, bty_f1lower))
```

As expected, the relationship is quite strong---after all, the average score is calculated using the individual scores. You can actually look at the relationships between all beauty variables (columns 13 through 19) using the following command:

```{r bty-rels}
evals %>%
  select(contains("bty")) %>%
  ggpairs()
```

These variables are collinear (correlated), and adding more than one of these variables to the model would not add much value to the model. In this application and with these highly-correlated predictors, it is reasonable to use the average beauty score as the single representative of these variables.

In order to see if beauty is still a significant predictor of professor score after you've accounted for the professor's gender, you can add the gender term into the model.

```{r scatter-score-bty_avg_pic-color}
m_bty_gen <- lm(score ~ bty_avg + gender, data = evals)
summary(m_bty_gen)
```

7.  P-values and parameter estimates should only be trusted if the
    conditions for the regression are reasonable. Verify that the conditions
    for this model are reasonable using diagnostic plots.

**Insert your answer here**
```{r}
par(mfrow = c(2, 2))
plot(m_bty_gen)
```

Check these diagnostic plots:

Residuals vs Fitted: No clear pattern → linearity OK
Q-Q Plot: Points follow line → normality OK
Scale-Location: Horizontal line → constant variance OK
Residuals vs Leverage: No extreme outliers → no influential points
Conclusion: Regression conditions are reasonably met.


8.  Is `bty_avg` still a significant predictor of `score`? Has the addition
    of `gender` to the model changed the parameter estimate for `bty_avg`?
    
**Insert your answer here**

Based on the two model outputs:

1. Is bty_avg still a significant predictor?
Yes. Its p-value in the new model is 6.48e-06, which is even more significant than before (5.08e-05).

2. Has the parameter estimate for bty_avg changed?

Yes. The coefficient increased from 0.0667 to 0.074.

**Interpretation:**
When we control for gender, the estimated effect of beauty on score becomes slightly stronger. A one-unit increase in bty_avg is now associated with a 0.074 point increase in score (compared to 0.067 before). 



Note that the estimate for `gender` is now called `gendermale`. You'll see this name change whenever you introduce a categorical variable. The reason is that R recodes `gender` from having the values of `male` and `female` to being an indicator variable called `gendermale` that takes a value of $0$ for female professors and a value of $1$ for male professors. (Such variables are often referred to as "dummy" variables.)

As a result, for female professors, the parameter estimate is multiplied by zero, leaving the intercept and slope form familiar from simple regression.

\[
  \begin{aligned}
\widehat{score} &= \hat{\beta}_0 + \hat{\beta}_1 \times bty\_avg + \hat{\beta}_2 \times (0) \\
&= \hat{\beta}_0 + \hat{\beta}_1 \times bty\_avg\end{aligned}
\]

<!-- We can plot this line and the line corresponding to those with color pictures
with the following  -->
<!-- custom function. -->

```{r twoLines}
ggplot(data = evals, aes(x = bty_avg, y = score, color = pic_color)) +
 geom_smooth(method = "lm", formula = y ~ x, se = FALSE)
```

9.  What is the equation of the line corresponding to those with color pictures? 
    (*Hint:* For those with color pictures, the parameter estimate is multiplied
    by 1.) For two professors who received the same beauty rating, which color 
    picture tends to have the higher course evaluation score?

**Insert your answer here**

```{r}
m_bty_pic <- lm(score ~ bty_avg + pic_color, data = evals)
summary(m_bty_pic)
```

Based on this model output:

The variable is pic_color, and the output shows a parameter for pic_colorcolor. This means:

Reference Group (coded as 0): pic_color = "black&white"
Comparison Group (coded as 1): pic_color = "color"

**1. Equation for color pictures:**
Black & white pictures: score = 4.06318 + 0.05548 × bty_avg

Color pictures: score = 4.06318 + 0.05548 × bty_avg - 0.16059

**2. Comparison:**
For two professors with the same beauty rating, the one with the black & white picture tends to have the higher evaluation score. The color picture coefficient is negative (-0.16)

Answer: The professor with the black & white picture tends to have the higher score. The model predicts that, for the same level of beauty, a color picture is associated with a significantly lower evaluation.

The decision to call the indicator variable `gendermale` instead of `genderfemale` has no deeper meaning. R simply codes the category that comes first alphabetically as a $0$. (You can change the reference level of a categorical variable, which is the level that is coded as a 0, using the`relevel()` function. Use `?relevel` to learn more.)

```{r}
?relevel
```

10. Create a new model called `m_bty_rank` with `gender` removed and `rank` 
    added in. How does R appear to handle categorical variables that have more 
    than two levels? Note that the rank variable has three levels: `teaching`, 
    `tenure track`, `tenured`.

**Insert your answer here**

```{r}
m_bty_rank <- lm(score ~ bty_avg + rank, data = evals)
summary(m_bty_rank)

```

The interpretation of the coefficients in multiple regression is slightly different from that of simple regression. The estimate for `bty_avg` reflects how much higher a group of professors is expected to score if they have a beauty rating that is one point higher *while holding all other variables constant*. In this case, that translates into considering only professors of the same rank with `bty_avg` scores that are one point apart.

## The search for the best model

We will start with a full model that predicts professor score based on rank, gender, ethnicity, language of the university where they got their degree, age, proportion of students that filled out evaluations, class size, course level, number of professors, number of credits, average beauty rating, outfit, and picture color.

11. Which variable would you expect to have the highest p-value in this model? 
    Why? *Hint:* Think about which variable would you expect to not have any 
    association with the professor score.

**Insert your answer here**

pic_outfit is most likely to be statistically irrelevant (truly random noise) in predicting teaching quality, making it the strongest candidate for a high p-value indicating no significant association.
This variable represents the type of random factor that should have no meaningful impact on teaching scores.

Let's run the model...

```{r m_full, tidy = FALSE}
m_full <- lm(score ~ rank + gender + ethnicity + language + age + cls_perc_eval 
             + cls_students + cls_level + cls_profs + cls_credits + bty_avg 
             + pic_outfit + pic_color, data = evals)
summary(m_full)
```

12. Check your suspicions from the previous exercise. Include the model output
    in your response.

**Insert your answer here**

  * Looking at the model output, the variable with the highest p-value is cls_profssingle with p = 0.77806.
  * While I initially suspected pic_outfit, cls_profssingle proves to be even more statistically irrelevant in predicting professor scores when controlling for all other variables

13. Interpret the coefficient associated with the ethnicity variable.
**Insert your answer here**

ethnicitynot minority  0.1234929  0.0786273   1.571  0.11698

Coefficient: ethnicitynot minority = 0.1235

Meaning: Professors who are "not minority" are predicted to have evaluation scores that are 0.123 points higher than minority professors, when all other variables in the model are held constant.

The p-value of 0.117 indicates this difference is not statistically significant at the conventional 0.05 level.



14. Drop the variable with the highest p-value and re-fit the model. Did the
    coefficients and significance of the other explanatory variables change?
    (One of the things that makes multiple regression interesting is that
    coefficient estimates depend on the other variables that are included in
    the model.) If not, what does this say about whether or not the dropped
    variable was collinear with the other explanatory variables?

**Insert your answer here**

Let's run the model excluding cls_profssingle...

```{r}
m_excluding_cls_profs <- lm(score ~ rank + gender + ethnicity + language + age + cls_perc_eval 
             + cls_students + cls_level  + cls_credits + bty_avg 
             + pic_outfit + pic_color, data = evals)
summary(m_excluding_cls_profs)
```

**Based on updated model results:**

1. Did coefficients and significance change?

No, the coefficients and significance levels changed very minimally.
P-values and significance patterns remained virtually identical

2. This demonstrates that cls_profssingle was NOT collinear with the other explanatory variables.

The minimal changes indicate cls_profssingle was an independent, irrelevant predictor that explained virtually no unique variance in scores.



15. Using backward-selection and p-value as the selection criterion,
    determine the best model. You do not need to show all steps in your
    answer, just the output for the final model. Also, write out the linear
    model for predicting score based on the final model you settle on.

**Insert your answer here**
```{r}
library(broom)
# Backward selection using p-value criterion
final_model <- step(m_full, direction = "backward", trace = 0)
#summary(final_model)
tidy(final_model)
```
```{r}
cat(paste("score =", paste(round(coef(final_model), 3), 
      names(coef(final_model)), collapse = " + ")))
```


The backward selection automatically removes non-significant variables, keeping only those that contribute meaningfully to predicting score


16. Verify that the conditions for this model are reasonable using diagnostic 
    plots.

**Insert your answer here**
```{r}
# Create diagnostic plots
par(mfrow = c(2, 2))
plot(final_model)
```

17. The original paper describes how these data were gathered by taking a
    sample of professors from the University of Texas at Austin and including 
    all courses that they have taught. Considering that each row represents a 
    course, could this new information have an impact on any of the conditions 
    of linear regression?

**Insert your answer here**

Yes, this information reveals a violation of the independence assumption.

**The Problem:**

  * Each row is a course, but multiple courses are taught by the same professor
  * Courses from the same professor are not independent observations - they share characteristics of that professor (teaching style, personality, etc.)
  * This creates clustered data where observations within the same professor are more similar to each other than to observations from different professors

**Impact:**
  1.  Standard errors may be underestimated
  2.  p-values may be too small
  3.  Confidence intervals may be too narrow
  4.  Risk of Type I errors (finding significant results that aren't actually significant)

18. Based on your final model, describe the characteristics of a professor and 
    course at University of Texas at Austin that would be associated with a high
    evaluation score.

**Insert your answer here**

Based on the final model, a professor with high evaluation scores would have these characteristics:

Professor Characteristics:

  * Male (+0.21 points vs. female) <br>
  * Higher average beauty rating (+0.05 points per 1-unit increase) <br>
  * English as native language (+0.20 points vs. non-English) <br>
  * Black & white photo (+0.19 points vs. color photo) <br>
  
**Course Characteristics:**


  * Higher percentage of students completing evaluations (+0.005 points per 1% increase)
  
  * One-credit course (+0.52 points vs. multi-credit courses)

**Key Takeaways:**
  * The strongest predictor is teaching a one-credit course, followed by being male and having a black & white photo. Beauty has a small but significant positive effect.
  
  * The model suggests both teaching-related factors (course type, evaluation response rate) and non-teaching biases (gender, appearance, language) influence ratings.


19. Would you be comfortable generalizing your conclusions to apply to professors
    generally (at any university)? Why or why not?
 
**Insert your answer here**    
No, I would not be comfortable generalizing these conclusions.

**Reasons:**

**Single University Sample:** Data comes only from University of Texas at Austin, which has its own unique institutional and student body characteristics.
**Regional/Cultural Bias:** Findings about beauty, gender, and photo color may reflect specific cultural biases at this university or region that don't apply universally.
**Selection Bias:** The sampling method (including all courses for selected professors) may not represent teaching assignments at other institutions.

The study identifies associations within this specific context, but these relationships likely vary across different universities and cultures.    




* * *
