---
title: "DATA 606 Data Project Proposal"
author: "Mehreen Ali Gillani"
output: html_document
date: "2025-11-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(1234)
library(tidyverse)
library(patchwork)
library(knitr)
library(kableExtra)
```



### Data Preparation

```{r}
library(readr)
#link to github data file 
flights_data_url <- 'https://raw.githubusercontent.com/mehreengillani/DATA606/refs/heads/main/flights_data.csv'


# read
flights <- read_csv(flights_data_url)

```

```{r}
# Data cleaning and preparation for domestic flights only
flights_clean <- flights %>%
  # Convert all column names to lowercase first
  rename_with(tolower) %>%
  filter(!is.na(passengers), !is.na(distance))

# Create variables using only existing categories in the data
flights_clean <- flights_clean %>%
  mutate(
    # Use only distance groups that actually exist in the data
    distance_cat = case_when(
      distance_group == 1 ~ "less than 500 miles",
      distance_group == 2 ~ "500-999 miles", 
      distance_group == 3 ~ "1000-1499 miles",
      distance_group == 4 ~ "1500-1999 miles",
      distance_group == 5 ~ "2000-2499 miles",
      distance_group == 6 ~ "2500-2999 miles",
      distance_group == 7 ~ "3000-3499 miles",
      distance_group == 8 ~ "3500-3999 miles",
      distance_group == 9 ~ "4000-4499 miles",
      distance_group == 10 ~ "4500-4999 miles",
      distance_group == 11 ~ "5000-5499 miles",
      distance_group == 12 ~ "5500-5999 miles",
      distance_group == 13 ~ "6000-6499 miles", 
      distance_group == 15 ~ "7000-7499 miles",
      TRUE ~ NA_character_  # Fixed: TRUE in uppercase
    ),
    
    # Use only carrier groups that actually exist in domestic data
    carrier_type = case_when(
      carrier_group == 1 ~ "regional carriers",
      carrier_group == 2 ~ "national carriers", 
      carrier_group == 3 ~ "major carriers",
      TRUE ~ NA_character_  # Fixed: TRUE in uppercase
    ),
    
    # Create month names for better visualization
    month_name = factor(month, 
                       levels = 1:7,
                       labels = c("jan", "feb", "mar", "apr", "may", "jun", "jul")),
    
    # Create route identifier
    route = paste(origin, dest, sep = " - "),
    
    # Calculate simple load metrics
    load_factor = passengers / 180
  )

# Remove rows with NA in critical categorical variables
flights_clean <- flights_clean %>%
  filter(!is.na(distance_cat), !is.na(carrier_type))

# Factor the categorical variables with only existing levels
existing_distance_cats <- c("less than 500 miles", "500-999 miles", "1000-1499 miles",
                           "1500-1999 miles", "2000-2499 miles", "2500-2999 miles",
                           "3000-3499 miles", "3500-3999 miles", "4000-4499 miles",
                           "4500-4999 miles", "5000-5499 miles", "5500-5999 miles",
                           "6000-6499 miles", "7000-7499 miles")

existing_carrier_cats <- c("major carriers", "national carriers", "regional carriers")

flights_clean$distance_cat <- factor(flights_clean$distance_cat,
                                    levels = existing_distance_cats)

flights_clean$carrier_type <- factor(flights_clean$carrier_type,
                                    levels = existing_carrier_cats)

# For analysis, create simplified distance categories based on actual data distribution
flights_clean <- flights_clean %>%
  mutate(
    distance_simple = case_when(
      distance_group %in% 1:2 ~ "short (<1000 mi)",
      distance_group %in% 3:5 ~ "medium (1000-2499 mi)", 
      distance_group %in% 6:10 ~ "long (2500-4999 mi)",
      distance_group >= 11 ~ "very long (5000+ mi)"
    ),
    distance_simple = factor(distance_simple,
                            levels = c("short (<1000 mi)", "medium (1000-2499 mi)",
                                      "long (2500-4999 mi)", "very long (5000+ mi)"))
  )




kable(head(flights_clean)) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(width = "100%", height = "400px")
```


### Research Questions

**1. Seasonal Flight Operations:** How do flight frequencies vary across months and carrier types?
Hypothesis: Summer months (June-July) will have the highest flight frequencies, with major carriers showing the most significant seasonal increases.
  * Response: flight_count (Numerical)
  * Explanatory: carrier_type (Categorical), month (Categorical) <br>
  
**2. Carrier Operational Strategies:** How do flight frequency patterns differ between carrier types?
Hypothesis: Major carriers maintain more consistent flight frequencies year-round, while regional carriers show greater seasonal variation.
  * Response: flight_count (Numerical)
  * Explanatory: carrier_type (Categorical), month (Categorical) <br>
  
  
**3. Market Dominance:** Which carriers dominate specific months in terms of flight operations?
Hypothesis: A few major carriers dominate flight operations across all months, with seasonal carriers emerging in peak months.
  * Response: flight_count (Numerical), market_share (Numerical)
  * Explanatory: unique_carrier_name (Categorical), month (Categorical)

####  Cases

What are the cases, and how many are there?

Each case represents a domestic flight segment operated by a U.S. carrier between two airports in a specific time period. There are 144875 observations in the cleaned dataset.

### Data Collection

Data is collected by the Bureau of Transportation Statistics (BTS) through mandatory reporting by all U.S. certified air carriers. Airlines submit Form 41 Traffic T-100 data, which includes detailed information about their scheduled domestic operations for 1-7 months in 2025, including passenger counts, freight, mail, and flight distances.

### Type of Study

What type of study is this (observational/experiment)?

This is an observational study.

### Data Source
data is downloaded from: https://www.transtats.bts.gov/DL_SelectFields.aspx?gnoyr_VQ=FIL&QO_fu146_anzr=Nv4+Pn44vr45
U.S. Department of Transportation, Bureau of Transportation Statistics, T-100 Domestic Segment Data.

Response

What is the response variable, and what type is it (numerical/categorical)?

The response variable is passengers (number of passengers on the flight segment) and is numerical.

Explanatory

What is the explanatory variable, and what type is it (numerical/categorical)?

The primary explanatory variable is distance (flight distance in miles) and is numerical. Additional exploratory variables include month (categorical) and carrier_type (categorical).

### Relevant Summary Statistics

Provide summary statistics relevant to your research question.

```{r}
# Summary statistics for main variable passengers
summary(flights_clean$passengers)

```
```{r}
# Summary statistics for distance
summary(flights_clean$distance)
```
```{r}
# Correlation between distance and passengers
cor(flights_clean$distance, flights_clean$passengers, use = "complete.obs")
```


```{r}
# Summary by distance categories
table(flights_clean$distance_cat, useNA = 'ifany')
```

```{r}
prop.table(table(flights_clean$distance_cat, useNA = 'ifany')) * 100
```


```{r}
library(psych)

# Passenger statistics by distance category
describeBy(flights_clean$passengers, group = flights_clean$distance_cat, mat = TRUE)

```
### Visualizations
```{r}
# Visualizations
ggplot(flights_clean, aes(x = passengers)) + 
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7) +
  labs(title = "Distribution of passengers per Flight Segment",
       x = "Number of passengers", y = "Frequency") +
  theme_minimal()
```

```{r}
ggplot(flights_clean, aes(x = distance)) + 
  geom_histogram(bins = 30, fill = "lightgreen", alpha = 0.7) +
  labs(title = "Distribution of Flight distances",
       x = "distance (miles)", y = "Frequency") +
  theme_minimal()

```

####  Flight distance vs No. of passengers
```{r}

# Scatter plot with transparency to handle overlapping points
p_main <- ggplot(flights_clean, aes(x = distance, y = passengers)) +
  geom_point(alpha = 0.1, size = 0.5, color = "darkblue") +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(title = "Flight distance vs. Number \n of passengers",
       subtitle = paste("Correlation:", round(cor(flights_clean$distance, flights_clean$passengers, 
                                                 use = "complete.obs"), 3)),
       x = "distance (miles)",
       y = "Number of passengers") +
  theme_minimal()

# Hexbin plot for better visualization of dense data
p_hex <- ggplot(flights_clean, aes(x = distance, y = passengers)) +
  geom_hex(bins = 50) +
  scale_fill_viridis_c(option = "plasma") +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(title = "distance vs passengers \n (Hexbin Density)",
       x = "distance (miles)",
       y = "Number of passengers",
       fill = "Count") +
  theme_minimal()

# Display both versions
p_main | p_hex
```

####  Total Flights by month

```{r}
# Calculate total flights by month
monthly_flights <- flights_clean %>%
  group_by(month) %>%
  summarise(
    total_flights = n(),
    .groups = 'drop'
  ) %>%
  arrange(desc(total_flights))

# Create month labels
month_labels <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul")

# Create proper month order while showing highest to lowest
monthly_flights_ordered <- monthly_flights %>%
  mutate(
    month_name = factor(month, 
                       levels = 1:7,
                       labels = month_labels[1:7]),
    rank = rank(-total_flights)
  ) %>%
  arrange(desc(total_flights))

# Create the plot
ggplot(monthly_flights_ordered, aes(x = reorder(month_name, -total_flights), 
                                   y = total_flights,
                                   fill = total_flights)) +
  geom_col(alpha = 0.8) +
  geom_text(aes(label = paste0(format(total_flights, big.mark = ","), 
                              "\n(", rank, ")")), 
            vjust = -0.08, size = 1.9, fontface = "bold") +
  scale_fill_gradient(low = "lightblue", high = "darkblue", 
                     name = "Total Flights") +
  labs(title = "monthly Flight Volume (January - July)",
       subtitle = "Ranked from highest to lowest, with rank numbers in parentheses",
       x = "month",
       y = "Total Number of Flights") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 10),
        plot.subtitle = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 0, hjust = 0.5))

```

```{r}
# Correlation between distance and month
cor(monthly_flights$month, monthly_flights$total_flights, use = "complete.obs")
```

There is a strong correlation between month and total flights

####  Passenger by carrier

```{r}

# Top 10 carriers by average passengers
top_carriers <- flights_clean %>%
  group_by(unique_carrier_name) %>%
  summarise(avg_passengers = mean(passengers, na.rm = TRUE),
            n_flights = n()) %>%
  filter(n_flights > 100) %>%  # Only carriers with sufficient data
  arrange(desc(avg_passengers)) %>%
  head(10)

ggplot(top_carriers, aes(x = reorder(unique_carrier_name, avg_passengers), 
                        y = avg_passengers, 
                        fill = avg_passengers)) +
  geom_col() +
  coord_flip() +
  labs(title = "Top 10 Carriers by Average passengers per Flight",
       x = "Carrier",
       y = "Average Number of passengers") +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  theme_minimal()
```

